---
title: "Storm Data"
output: html_document
keep_md: yes
pdf_document: default
word_document: default
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, fig.width = 12)
```

## Synopsis
The purpose of this analysis is to help answer the following questions:
* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?


## Data Processing

[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

```{r load_and_preprocess, cache=FALSE}
# loading ALL required libraries
library(tools)
library(ggplot2)
library(scales)
library(plyr)
library(stringr)
library(R.utils)
library(sqldf)

# file information (downloaded on 4/21/2015 3:20 PM CDT)
filelink <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filename <- "repdata_data_StormData.csv.bz2"
filechecksum <- "df4aa61fff89427db6b7f7b1113b5553"  
extractedfile <- "Extracted Data/repdata_data_StormData.csv"
extractedfilechecksum <- "33ab0bd27d935eeefef0dd7300f800af"

# check to see if file is missing or checksum has changed 
if (!file.exists(filename)|md5sum(filename) != filechecksum) {
        # download the file.
        download.file(filelink, filename)
        }

# check to see if the extracted file exists or the checksum doesn't match
if (!file.exists(extractedfile) | md5sum(extractedfile) != extractedfilechecksum) {
        # extract file
        bunzip2(filename = filename, destname = extractedfile)
        }

# read in data file
# sd <- read.csv(bzfile(filename))
#sd <- read.csv(extractedfile)        


# functions to normalize event types
clean.EVTYPE <- function(x, debug=FALSE){
        library(stringr)
        valid.EVTYPE <- c("ASTRONOMICAL LOW TIDE",
                          "AVALANCHE",
                          "BLIZZARD",
                          "COASTAL FLOOD",
                          "COLD/WIND CHILL",
                          "DEBRIS FLOW",
                          "DENSE FOG",
                          "DENSE SMOKE",
                          "DROUGHT",
                          "DUST DEVIL",
                          "DUST STORM",
                          "EXCESSIVE HEAT",
                          "EXTREME COLD/WIND CHILL",
                          "FLASH FLOOD",
                          "FLOOD",
                          "FROST/FREEZE",
                          "FUNNEL CLOUD",
                          "FREEZING FOG",
                          "HAIL",
                          "HEAT",
                          "HEAVY RAIN",
                          "HEAVY SNOW",
                          "HIGH SURF",
                          "HIGH WIND",
                          "HURRICANE (TYPHOON)",
                          "ICE STORM",
                          "LAKE-EFFECT SNOW",
                          "LAKESHORE FLOOD",
                          "LIGHTNING",
                          "MARINE HAIL",
                          "MARINE HIGH WIND",
                          "MARINE STRONG WIND",
                          "MARINE THUNDERSTORM WIND",
                          "RIP CURRENT",
                          "SEICHE",
                          "SLEET",
                          "STORM SURGE/TIDE",
                          "STRONG WIND",
                          "THUNDERSTORM WIND",
                          "TORNADO",
                          "TROPICAL DEPRESSION",
                          "TROPICAL STORM",
                          "TSUNAMI",
                          "VOLCANIC ASH",
                          "WATERSPOUT",
                          "WILDFIRE",
                          "WINTER STORM",
                          "WINTER WEATHER")

        # remove spaces before and after
        new.EVTYPE <- str_trim(x, side="both")
        
        # change to upper case
        new.EVTYPE <- toupper(new.EVTYPE)
        
        # remove misc punctuation
        new.EVTYPE <- sub("\\.$", "", new.EVTYPE)
        new.EVTYPE <- sub(";$", "", new.EVTYPE)
        new.EVTYPE <- sub("\\(", " ", new.EVTYPE) # note: this will break "HURRICANE (TYPHOON)", but we fix it later
        new.EVTYPE <- sub(")$", "", new.EVTYPE)   # note: this will break "HURRICANE (TYPHOON)", but we fix it later
        
        
        # remove all double spaces within value
        while(length(grep("  ", new.EVTYPE)) > 0){new.EVTYPE <- sub("  ", " ", new.EVTYPE)}
        
        # remove spaces before and after slashes and stanardize them
        new.EVTYPE <- sub("-", "/", new.EVTYPE)
        new.EVTYPE <- sub(";", "/", new.EVTYPE)
        new.EVTYPE <- sub("\\\\", "/", new.EVTYPE)
        new.EVTYPE <- sub("/ ", "/", new.EVTYPE)
        new.EVTYPE <- sub(" /", "/", new.EVTYPE)
        
        # replace abbreviations
        new.EVTYPE <- sub("TSTM ", "THUNDERSTORM ", new.EVTYPE)
        new.EVTYPE <- sub("SML ", "SMALL ", new.EVTYPE)
        new.EVTYPE <- sub("FLD", "FLOOD", new.EVTYPE)
        
        # fix typos
        new.EVTYPE <- sub("^AVALANCE$", "AVALANCHE", new.EVTYPE)
        new.EVTYPE <- sub("THUNDERTORM", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("^THUNDERSTORMW$", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("TUNDERSTORM", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("THUNERSTORM", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("THUNDERSTROM", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("THUNDERESTORM", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("THUNDEERSTORM", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("THUDERSTORM", "THUNDERSTORM", new.EVTYPE)
        new.EVTYPE <- sub("WINS", "WINDS", new.EVTYPE)
        new.EVTYPE <- sub("^LIGHTING$", "LIGHTNING", new.EVTYPE)
        new.EVTYPE <- sub("^TORNDAO", "TORNDAO", new.EVTYPE)
 
        # remove descriptors
        new.EVTYPE <- sub("^MAJOR ", "", new.EVTYPE)
        new.EVTYPE <- sub("^MINOR ", "", new.EVTYPE)
        new.EVTYPE <- sub("^SEVERE ", "", new.EVTYPE)
        new.EVTYPE <- sub("^SMALL ", "", new.EVTYPE)
        new.EVTYPE <- sub("^URBAN ", "", new.EVTYPE)
        new.EVTYPE <- sub("^URBAN/SMALL ", "", new.EVTYPE)
        new.EVTYPE <- sub("^STREAM ", "", new.EVTYPE)
        
        new.EVTYPE <- sub("^EROSION/", "", new.EVTYPE)        
        
        # remove plurals and verbs
        new.EVTYPE <- sub("CURRENTS$", "CURRENT", new.EVTYPE)
        new.EVTYPE <- sub("WINDS$", "WIND", new.EVTYPE)
        new.EVTYPE <- sub("STORMS", "STORM", new.EVTYPE)
        new.EVTYPE <- sub("FLOODING", "FLOOD", new.EVTYPE)
        new.EVTYPE <- sub("FLOODS", "FLOOD", new.EVTYPE)
        new.EVTYPE <- sub("RAINS", "RAIN", new.EVTYPE)
        new.EVTYPE <- sub("WINDS", "WIND", new.EVTYPE)
        new.EVTYPE <- sub("FIRES", "FIRE", new.EVTYPE)
        new.EVTYPE <- sub("TORNADOES", "TORNADO", new.EVTYPE)

        # coerce into standard event types
        # new.EVTYPE <- sub("^$", "", new.EVTYPE)
        
        ## BLIZZARD
        new.EVTYPE <- sub("^HEAVY SNOW AND HIGH WINDS$", "BLIZZARD", new.EVTYPE)
        
        ## COASTAL FLOOD
        new.EVTYPE <- sub("^COASTAL FLOOD.*$", "COASTAL FLOOD", new.EVTYPE)
        
        ## COLD/WIND CHILL
        new.EVTYPE <- sub("^COLD$", "COLD/WIND CHILL", new.EVTYPE)
        
        ## DENSE FOG
        new.EVTYPE <- sub("^FOG$", "DENSE FOG", new.EVTYPE)
        
        ## EXCESSIVE HEAT
        new.EVTYPE <- sub("^EXTREME HEAT$", "EXCESSIVE HEAT", new.EVTYPE)
        
        ## EXTREME COLD/WIND CHILL
        new.EVTYPE <- sub("^EXTREME COLD.*$", "EXTREME COLD/WIND CHILL", new.EVTYPE)
        new.EVTYPE <- sub("^RECORD COLD.*$", "EXTREME COLD/WIND CHILL", new.EVTYPE)
        
        ## FLASH FLOOD
        new.EVTYPE <- sub("^FLASH FLOOD.*$", "FLASH FLOOD", new.EVTYPE)
        
        ## FLOOD
        new.EVTYPE <- sub("^RIVER FLOOD$", "FLOOD", new.EVTYPE)
        new.EVTYPE <- sub("^SNOWMELT FLOOD$", "FLOOD", new.EVTYPE)
        new.EVTYPE <- sub("^ICE JAM FLOOD$", "FLOOD", new.EVTYPE)
        new.EVTYPE <- sub("^FLOOD.*$", "FLOOD", new.EVTYPE)

        ## HAIL
        new.EVTYPE <- sub("^HAIL.*$", "HAIL", new.EVTYPE)
        
        ## HEAT
        new.EVTYPE <- sub("^HEAT WAVE$", "HEAT", new.EVTYPE)
        
        ## HEAVY RAIN
        new.EVTYPE <- sub("^HEAVY RAIN.*$", "HEAVY RAIN", new.EVTYPE)
        new.EVTYPE <- sub("^RAIN$", "HEAVY RAIN", new.EVTYPE)
        
        ## HEAVY SNOW
        new.EVTYPE <- sub("^HEAVY SNOW.*$", "HEAVY SNOW", new.EVTYPE)
        new.EVTYPE <- sub("^HEAVY SNOW.*$", "HEAVY SNOW", new.EVTYPE)
        new.EVTYPE <- sub("^SNOW( |/).*$", "WINTER WEATHER", new.EVTYPE)
        new.EVTYPE <- sub("^SNOW( |/).*$", "WINTER WEATHER", new.EVTYPE)
        new.EVTYPE <- sub("^SNOW$", "WINTER WEATHER", new.EVTYPE)
        
        ## HIGH SURF
        new.EVTYPE <- sub("^HEAVY SURF/HIGH SURF$", "HIGH SURF", new.EVTYPE)
        new.EVTYPE <- sub("^HEAVY SURF$", "HIGH SURF", new.EVTYPE)
        
        ## HIGH WIND
        new.EVTYPE <- sub("^HIGH WIND.*$", "HIGH WIND", new.EVTYPE)
 
        ## HURRICANE (TYPHOON)
        new.EVTYPE <- sub("^HURRICANE .*$", "HURRICANE (TYPHOON)", new.EVTYPE)
        new.EVTYPE <- sub("^HURRICANE/TYPHOON$", "HURRICANE (TYPHOON)", new.EVTYPE)
        new.EVTYPE <- sub("^HURRICANE$", "HURRICANE (TYPHOON)", new.EVTYPE)
        new.EVTYPE <- sub("^TYPHOON$", "HURRICANE (TYPHOON)", new.EVTYPE)

        ## ICE STORM
        new.EVTYPE <- sub("^ICE$", "ICE STORM", new.EVTYPE)
        
        ## LIGHTNING
        new.EVTYPE <- sub("^LIGHTNING.*$", "LIGHTNING", new.EVTYPE)
        
        ## STORM SURGE/TIDE
        new.EVTYPE <- sub("^STORM SURGE$", "STORM SURGE/TIDE", new.EVTYPE)
        
        ## STRONG WIND
        new.EVTYPE <- sub("^GUSTY WIND$", "STRONG WIND", new.EVTYPE)
        new.EVTYPE <- sub("^WIND$", "STRONG WIND", new.EVTYPE)
        
        ## THUNDERSTORM WIND
        new.EVTYPE <- sub("^THUNDERSTORM$", "THUNDERSTORM WIND", new.EVTYPE)
        new.EVTYPE <- sub("^THUNDERSTORM WIND.*$", "THUNDERSTORM WIND", new.EVTYPE)
        new.EVTYPE <- sub("^TSTMW$", "THUNDERSTORM WIND", new.EVTYPE)
        new.EVTYPE <- sub("^THUNDERSTORMWINDS$", "THUNDERSTORM WIND", new.EVTYPE)
 
        ## TORNADO
        new.EVTYPE <- sub("^WATERSPOUT/TORNADO$", "TORNADO", new.EVTYPE)
        new.EVTYPE <- sub("^TORNADO.*$", "TORNADO", new.EVTYPE)
        
        ## TROPICAL STORM
        new.EVTYPE <- sub("^TROPICAL STORM.*$", "TROPICAL STORM", new.EVTYPE)

        ## WILDFIRE
        new.EVTYPE <- sub("^WILD FIRE$", "WILDFIRE", new.EVTYPE)
        new.EVTYPE <- sub("^WILD/FOREST FIRE$", "WILDFIRE", new.EVTYPE)
        new.EVTYPE <- sub("^FOREST FIRE$", "WILDFIRE", new.EVTYPE)
        
        ## WINTER STORM
        new.EVTYPE <- sub("^WINTER STORM.*$", "WINTER STORM", new.EVTYPE)
        
        ## WINTER WEATHER
        new.EVTYPE <- sub("^WINTER WEATHER MIX$", "WINTER WEATHER", new.EVTYPE)
        new.EVTYPE <- sub("^WINTER WEATHER/MIX$", "WINTER WEATHER", new.EVTYPE)
        new.EVTYPE <- sub("^WINTRY MIX$", "WINTER WEATHER", new.EVTYPE)
        
        
        
        ## anything that isn't matched gets lumped together as "UNKNOWN/OTHER"
        if(!debug){
                missing.EVTYPE <- !(new.EVTYPE %in% valid.EVTYPE)
                
                for (i in 1:length(new.EVTYPE)) {
                        if(missing.EVTYPE[i] == TRUE) new.EVTYPE[i] <- "UNKNOWN/OTHER"
                        }

                }
     
        return(new.EVTYPE)
        
        }

```

## Results
### Health Impact

```{r health_impact}
# summarize the data by event type
sdh <- ddply(sd, c("EVTYPE"), summarize, INJ_FAT = sum(INJURIES + FATALITIES))

# remove any items where no one was injured or killed
sdh <- subset(sdh, sdh$INJ_FAT != 0)

# clean up remaining event types
sdh$EVTYPE <- clean.EVTYPE(sdh$EVTYPE)

# resummarize on cleaned up event types
sdh <- ddply(sdh, c("EVTYPE"), summarize, INJ_FAT = sum(INJ_FAT))

# grab top 10 event types
sdh_10 <- head(sdh[with(sdh, order(-INJ_FAT, EVTYPE)),], 10)

# plot them
ggplot(sdh_10, 
       aes(reorder(x=factor(EVTYPE), 
                   -INJ_FAT
                   ),
           y=INJ_FAT
           )
       ) +
        geom_bar(stat="identity") +
        theme(axis.text.x = element_text(angle= 90,
                                         hjust = 1)
              ) +
        labs(title = "Top 10 Weather Events Causing Injuries and/or Fatalities",
             x = "Weather Event Type",
             y = "Number of Injuries and/or Fatalities")
```

### Ecomnomic Impact
```{r economic_impact}
# start pulling out just the needed columns
sde <- subset(sd, select=c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))

# normalize the multipliers
sde$PROPDMGEXP <- toupper(sde$PROPDMGEXP)
sde$CROPDMGEXP <- toupper(sde$CROPDMGEXP)

# these are the only defined multipliers
valid_EXP <- data.frame(cbind(EXP_values = c("K", "M", "B"), EXP_mulitpliers = c(10^3, 10^6, 10^9)))

# this query pulls in the multiplier value
sql.1 <- "select EVTYPE, PROPDMG, PROPDMGEXP, COALESCE(PEXP.EXP_mulitpliers, 0) PEXP_mulitplier, CROPDMG, CROPDMGEXP, COALESCE(CEXP.EXP_mulitpliers, 0) CEXP_mulitplier"
sql.1 <- paste(sql.1, " from sde left outer join valid_EXP PEXP on sde.PROPDMGEXP = PEXP.EXP_values")
sql.1 <- paste(sql.1, " left outer join valid_EXP CEXP on sde.CROPDMGEXP = CEXP.EXP_values")
sde <- sqldf(sql.1)

# calculate the total damage
sql.2 <- "select EVTYPE, ((PROPDMG * PEXP_mulitplier) + (CROPDMG * CEXP_mulitplier)) TOTAL_DMG from sde"
sde <- sqldf(sql.2)
sde <- subset(sde, sde$TOTAL_DMG > 0)

# clean up remaining event types
sde$EVTYPE <- clean.EVTYPE(sde$EVTYPE)

# summarize the data by the normalized event types
sde <- ddply(sde, c("EVTYPE"), summarize, TOTAL_DMG = sum(TOTAL_DMG))

# grab top 10 event types
sde_10 <- head(sde[with(sde, order(-TOTAL_DMG, EVTYPE)),], 10)

# plot them
ggplot(sde_10, 
       aes(reorder(x=factor(EVTYPE), 
                   -TOTAL_DMG
                   ),
           y=TOTAL_DMG/10^6
           )
       ) +
        geom_bar(stat="identity") +
        theme(axis.text.x = element_text(angle= 90,
                                         hjust = 1)
              ) +
        labs(title = "Top 10 Weather Events Causing Property and/or Crop Damage",
             x = "Weather Event Type",
             y = "Property and/or Crop Damage (in millions of dollars")

```



## Results







